Probabilistic bioregion models suggest seabird assemblages are linked to the major boundary current off eastern Australia

### 2023-06-13 docs/ms-draft.Rmd
runing figs again and sorting plots for ms; few adjustments on the model fitting/results scripts to export ggplot objs as well as constantly updating figs-patchwork script; in source script I noticed it was missing one require() inside the prob-maps plot function so I added it

---
title: "Probabilistic bioregion models suggest seabird assemblages are linked to the boundary current off eastern Australia"
# "Probabilistic bioregion models suggest seabird assemblages are linked to large scale oceanography off eastern Australia"
# author: Nicholas W Daudt
# date: '`r format(Sys.Date())`'
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [ASG]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - ASG: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species at large scale, probabilistically, and related them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used seabird at-sea ship-based survey data collected over 37° latitude and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and counts). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within it but different from other regions. These groups also vary according to covariates, on which we included oceanographic static and climatic, as well as physiographic layers.
  **Results**: All models suggested two large scale assemblages ('northern' and 'southern' assemblages), except for the abundance-autumn model which indicated three groups. Model results seem to agree in a biogeographic barrier at ~32°S, where the East Australian Current mostly dissipates to the east and loses force. Furthermore, sea surface temperature and, in one case, sea surface salinity, were selected in all final models, further indicating close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and counts resulted in similar spatial and species profile patterns.
  **Conclusion**: Two seabird assemblages were clearly identified from RCP models, suggesting the persistence of these groups at temporal and spatial scales. Given the rapid change that the East Australian Current is experiencing, and its importance on driving seabirds' distributions, the methods applied in our study could be replicated in future analyses to assess possible spatial/temporal changes in seabird assemblages, including on its structure.
output: 
  # html_document: default
  # word_document:
  #   reference_docx: "../docs/docx-template.docx"
  # officedown::rdocx_document:
  #   reference_docx: "../docs/docx-template.docx"
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

@@

```{r fig2-point-pred, fig.cap = "**FIGURE 2.** Spatialized point predictions for seabird RCP models, by season, based on presence/absence data (top row) and abundance data (bottom row) off eastern Australia. Each cell was assigned to the RCP group with highest probability (i.e. hard-classes).", fig.align = "center", echo = FALSE}

@@

```{r fig3-prob-maps, fig.cap = "**FIGURE 3.** Predicted probabilities of RCP membership in in spring presence/absence (a) and abundance (b) models.", fig.align = "center", echo = FALSE}

@@

>>>>

---
title: "Probabilistic bioregion models suggest seabird assemblages are linked to the major boundary current off eastern Australia"
# "Probabilistic bioregion models suggest seabird assemblages are linked to large scale oceanography off eastern Australia"
# author: Nicholas W Daudt
# date: '`r format(Sys.Date())`'
author:
  - Nicholas W. Daudt:
      email: nicholaswdaudt@gmail.com
      institute: [MarSciUO]
      correspondence: true
  - Eric J. Woehler:
      institute: [ASG]
      correspondence: false
  - Matthew R. Schofield:
      institute: [MathStatUO]
      correspondence: false
  - Robert O. Smith:
      institute: [MarSciUO]
      correspondence: false
  - Leandro Bugoni:
      institute: [FURG]
      correspondence: false
  - William J. Rayment:
      institute: [MarSciUO]
      correspondence: false
institute:
  - MarSciUO: Department of Marine Science, University of Otago, Aotearoa New Zealand
  - ASG: Australasian Seabird Group, BirdLife Australia, Hobart, Australia
  - MathStatUO: Department of Mathematics and Statistics, University of Otago, Aotearoa New Zealand
  - FURG: Institute of Biological Sciences, Universidade Federal do Rio Grande--FURG, Rio Grande, RS, Brazil
abstract: |
  **Aim**: To identify assemblages of seabird species at large scale, probabilistically, and related them to environmental conditions off eastern Australia, a world seabird hotspot.
  **Methods**: We used seabird at-sea ship-based survey data collected over 37° latitude and fitted Region of Common Profile (RCP) mixture models for each season and type of data (presence-absence and counts). RCP groups are defined as a set of regions where the probability of encountering a specific species profile is constant within it but different from other regions. These groups also vary according to covariates, on which we included oceanographic static and climatic, as well as physiographic layers.
  **Results**: All models suggested two large scale assemblages ('northern' and 'southern' assemblages), except for the abundance-autumn model which indicated three groups. Model results seem to agree in a biogeographic barrier at ~32°S, where the East Australian Current mostly dissipates to the east and loses force. Furthermore, sea surface temperature and, in one case, sea surface salinity, were selected in all final models, further indicating close relationship between seabird assemblages and water masses. Results from both data types, presence-absence and counts resulted in similar spatial and species profile patterns.
  **Conclusion**: Two seabird assemblages were clearly identified from RCP models, suggesting the persistence of these groups at temporal and spatial scales. Given the rapid change that the East Australian Current is experiencing, and its importance on driving seabirds' distributions, the methods applied in our study could be replicated in future analyses to assess possible spatial/temporal changes in seabird assemblages, including on its structure.
output: 
  # html_document: default
  # word_document:
  #   reference_docx: "../docs/docx-template.docx"
  # officedown::rdocx_document:
  #   reference_docx: "../docs/docx-template.docx"
  # rticles::sim_article:
  #   keep_tex: TRUE
  bookdown::word_document2:
    pandoc_args:
      - --lua-filter=../docs/scholarly-metadata.lua
      - --lua-filter=../docs/author-info-blocks.lua
      - --reference-doc=../docs/docx-template.docx 
bibliography: ../references.bib
csl: https://raw.githubusercontent.com/citation-style-language/styles/master/apa.csl
---

@@

```{r fig2-point-pred, fig.cap = "**FIGURE 2.** Spatialized point predictions (hard-classes) for seabird Region of Common Profile (RCP) models, by season, based on presence-absence (a) and abundance (b) models off eastern Australia. Each grid was assigned to the RCP group with highest membership probability.", fig.align = "center", echo = FALSE}

@@

```{r fig3-prob-maps, fig.cap = "**FIGURE 3.** Predicted probabilities of Region of Common Profiles membership for each grid, based on selected covariates (see Table 2) of spring presence-absence (a) and abundance (b) models for seabird assemblages off eastern Australia. Results for other seasons and models can be found in the Supplementary Material.", fig.align = "center", echo = FALSE}

@@

```{r fig5-spp-profiles, fig.cap = "**FIGURE 5.** Species profiles of seabird assembalges off eastern Australia given by spring presence-absence (a) and abundance (b) Region of Common Profile models. Note that values in (b) are log10 transformed. Results for other seasons and models can be found in the Supplementary Material.", fig.align = "center", echo = FALSE}

knitr::include_graphics("../results/Fig5-spp-profiles-spring.png")
```
